---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
  
## Loading and preprocessing the data
```{r getdata, echo = TRUE}
url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, "repdata-data-activity.zip", mode = "wb")
unzip("repdata-data-activity.zip")
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
```
  
## What is mean total number of steps taken per day?
```{r exploredata, echo = TRUE}
daily.steps <- tapply(data$steps, data$date, sum)
hist(daily.steps, main = "Total nuber of steps per day", xlab = "Date", ylab = "Total number of steps")
mean <- mean(daily.steps, na.rm = TRUE)
mean.string <- as.character(round(mean, digits = 2))
median <- median(daily.steps, na.rm = TRUE)
abline(v = mean, col = "red")
legend("topright", "mean", lty = c(1, 1), col = "red")
```
The average number of steps taken per day is `r mean.string` (red line) and the median is `r median`.  
  
## What is the average daily activity pattern?
```{r pattern, echo = TRUE}
int.steps <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
plot(as.numeric(names(int.steps)), int.steps, type = 'l', main = "Average number of steps per daily interval", xlab = "Interval", ylab = "Avera steps")
max <- max(int.steps)
max.string <- as.character(round(max, digits = 2))
int <- names(int.steps[which(int.steps == max)])
abline(v = int, col = "blue")
```
The daily observations show a peak at the interval `r int` (blue line) with a maximum of `r max.string` average number of steps.  
  
## Imputing missing values
```{r imputdata, echo = TRUE}
NA.count <- sum(is.na(data$steps))
data.adj <- data
missing.days <- sum(is.na(daily.steps))
for (i in 1: length(daily.steps)) {
    if (is.na(daily.steps[i])) {
        # substitute all NAs with the average value of the interval
        data.adj$steps[data.adj$date == names(daily.steps)[i]] <- int.steps
    }
}
```
The number of missing data is `r NA.count`. All missing are in data have been replaced by the average number of steps of the interval they belonged to.
```{r imputdata2, echo = TRUE}
daily.steps.adj <- tapply(data.adj$steps, data.adj$date, sum)
hist(daily.steps.adj, main = "Total nuber of steps per day", xlab = "Date", ylab = "Total number of steps")
mean.adj <- mean(daily.steps.adj)
mean.adj.string <- as.character(round(mean.adj, digits = 2))
median.adj <- median(daily.steps.adj)
median.adj.string <- as.character(round(median.adj, digits = 2))
```
The new average number of steps per day is `r mean.adj.string` and the new median is `r median.adj.string`. The mean has not changed, as we expected, however the central bin in new histogram increased by `r missing.days`. The reason is that all missing data belonged to `r missing.days` specific days having a total number of steps equal to the mean and the mean falls in the central bin. The median has suffered a small increase because it falls in one of the days having missing data and is equal to the mean (by construction).  

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays, echo = TRUE, fig.height = 8}
data.adj$week <- weekdays(data.adj$date)
data.adj$week[data.adj$week == "Saturday" | data.adj$week == "Sunday"] <- "weekend"
data.adj$week[data.adj$week != "weekend"] <- "weekday"
int.steps2 <- data.frame(with(data.adj, tapply(steps, list(interval, week), mean)))
with(int.steps2, {
    par(mfrow = c(2, 1))
    plot(rownames(int.steps2), int.steps2[, 1], type = "l", main = colnames(int.steps2)[1], xlab = "Interval", ylab = "Number of teps")
    plot(rownames(int.steps2), int.steps2[, 2], type = "l", main = colnames(int.steps2)[2], xlab = "Interval", ylab = "Number of teps")
})
```