---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
  
## Loading and preprocessing the data
```{r getdata, echo = TRUE}
url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, "repdata-data-activity.zip", mode = "wb")
unzip("repdata-data-activity.zip")
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
```
  
## What is mean total number of steps taken per day?
```{r exploredata, echo = TRUE}
daily.steps <- tapply(data$steps, data$date, sum)
barplot(daily.steps, main = "Total nuber of steps per day", xlab = "Date", ylab = "Total number of steps")
mean <- mean(daily.steps, na.rm = TRUE)
mean.string <- as.character(round(mean, digits = 2))
median <- median(daily.steps, na.rm = TRUE)
```
The average number of steps taken per day is `r mean.string` and the median is `r median`.  
  
## What is the average daily activity pattern?
```{r pattern, echo = TRUE}
int.steps <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
plot(as.numeric(names(int.steps)), int.steps, type = 'l', main = "Average number of steps per daily interval", xlab = "Interval", ylab = "Avera steps")
max <- max(int.steps)
max.string <- as.character(round(max, digits = 2))
int <- names(int.steps[which(int.steps == max)])
```
The daily observations show a peak around the interval `r int` with a maximum of `r max.string` average number of steps.  
  
## Imputing missing values
```{r imputdata, echo = TRUE}
NA.count <- sum(is.na(data$steps))
data.adj <- data
for (i in 1: length(daily.steps)) {
    if (is.na(daily.steps[i])) {
        # substitute all NAs with the average value of the interval
        data.adj$steps[data.adj$date == names(daily.steps)[i]] <- int.steps
    }
}
```
The number of missing data is `r NA.count`. All missing data have been replaced by the average number of steps of the interval they belonged to.
```{r imputdata2, echo = TRUE}
daily.steps.adj <- tapply(data.adj$steps, data.adj$date, sum)
barplot(daily.steps.adj, main = "Total nuber of steps per day", xlab = "Date", ylab = "Total number of steps")
mean.adj <- mean(daily.steps.adj)
mean.adj.string <- as.character(round(mean.adj, digits = 2))
median.adj <- median(daily.steps.adj)
median.adj.string <- as.character(round(median.adj, digits = 2))
```
The new average number of steps per day is `r mean.adj.string` and the new median is `r median.adj.string`. The mean has not changed, as we expected. The median has suffered a small increase because missing values were replaced with averages in decimal format, whereas steps are in units. The new median falls on one of the days ahving missing data and is equal by construction to the mean.
  
## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays, echo = TRUE, fig.height = 8}
data.adj$week <- weekdays(data.adj$date)
data.adj$week[data.adj$week == "Saturday" | data.adj$week == "Sunday"] <- "weekend"
data.adj$week[data.adj$week != "weekend"] <- "weekday"
int.steps2 <- data.frame(with(data.adj, tapply(steps, list(interval, week), mean)))
with(int.steps2, {
    par(mfrow = c(2, 1))
    plot(rownames(int.steps2), int.steps2[, 1], type = "l", main = colnames(int.steps2)[1], xlab = "Interval", ylab = "Number of teps")
    plot(rownames(int.steps2), int.steps2[, 2], type = "l", main = colnames(int.steps2)[2], xlab = "Interval", ylab = "Number of teps")
})
```
